Avoiding the Global Sort: \break A Faster Contour Tree Algorithm

### 2017-05-03 response.txt

Review #1:

The authors have updated the paper and fixed many issues. They have addressed my concerns. However, these additions have caused some new minor issues, and, in my opinion, still the the paper is not mature enough to be published as it is, as will be seen in the comments below. I will mention several more important points in the following, but I was unable to list all the typos.

Global issues:
1. Since the paper uses the results of CSA03 for merging a split tree with a merge tree to form a contour tree, it is crucial to have the same type of domains. So I suggest that the authors define their initial domain to be the triangulation of a ball, and again crucially, they should mention that the pieces that they break the domain into, i.e. extremum dominant complexes, are also triangulations of topological balls, so that they can always use the results of CSA03. This is the case since the pieces are formed by raining. Please verify this condition to use the theorems from CSA03. In doing so, the authors can update Theorem 5.4 to say that the outputs of cut are triangulated balls. Again, this is necessary since at each recursive step that the results of CSA03 are used the domains must be triangulated topological balls.
The current wording of definition of M is not good. When defined as in CSA03 to be a triangulated ball, then the simply-connectedness assumption is not needed and they can remove it.
--->We have fixed the wording in Theorem 1.1, 1.3, and Theorem 5.4. We have verified that this is consistent with all our previous wording and with the results of CSA03.
(OLD--->We explicitly did not want to triangulate the boundary faces, since we were worried about situations like a double walled hollow upside bowl (i.e. there is another maximum poking through the boundary facet).  However, I am not entirely sure that is possible given we are lifting the input complex one dimension higher?  At this point I am fine just doing what the reviewer says though.  Can you figure this out?
Also small side note, I added "except for pairs from the same boundary facet," to section 2, first paragraph, last sentence:
)

2. Make sure that all headings have the same uppercase lowercase convention, see for example 1.2, 4.4 etc.
--->All section and subsection heading now have the same lowercase convention, except for the leading word being capitalized.

3. I suggest that the authors go through the updated sections more carefully and correct any typos.
---> We went through the updated sections again as suggested to check for and correct typos.

Smaller issues, typos:
1. In the abstract please define n.
---> Defined.

2. In the previous work it seems that Vyalya should be Vyalyi.
--->Corrected

3. In the second par. of the Introduction, the sentence in the parenthesis is very strange indeed. They should remove it and correct the sentence before it. Not all critical points appear in the contour tree. They can simply say some of these critical points affect the connected components of the contour, and those correspond to vertices of the contour tree. Indeed the contour tree is a simple construct and explaining it does not need more complex definitions.
--->We fixed this paragraph.

4. In the third paragraph of the Introduction, the assumption on Morse degree 3 cannot be thought of as general position assumption. Please remove this. You might want to mention that it is possible to do "vertex unfolding" in the domain.
--->We removed this sentence.
(OLD--->I commented out the general position sentence and added "The degree assumption can be achieved via vertex unfolding, as discussed later,". Please take a look.)

5. In 2.1 it is perhaps good to number the remarks 1) and 2).
---> The remarks are now numbered.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Second review:
The changes and the reorganization of the paper by the authors are all accept-
able. There is a typo and a grammatical error (listed below) which need to be
fixed. I have two other comments, one a required inclusion and one a strong
suggestion.

a. The authors added to Theorem 1.1, the statement ”Let the Morse degree
of f at any point be at most 3.” They should also state this condition in
Theorem 1.3 and in the Abstract. While they note that assuming Morse
degree three ”is a standard assumption in the literature”, this is standard
only for algorithms constructing Morse complexes. I don’t believe that the
algorithms in either CSA03 or CLLR05 make or require this assumption.
This is a real difference so the assumption about Morse degree should be
in the abstract and Theorem 1.3. (The authors also should also add ”See
Definition 2.3 for a definition of Morse Degree” after using the term Morse
Degree in Theorem 1.1.)
--->We updated the abstract. We put a sentence in Theorem 1.3 referring to the conditions in Theorem 1.1.

b. The authors followed my suggestion of moving the join tree algorithm
earlier and adding a higher level description. However, I repeat my rec-
ommendation that they provide a simplicial complex illustration of the
algorithm in Section 4.2.2 The illustration in Figure 3 is good, but it does
not illustrate the details of the algorithm in 4.2.2. If an illustration is
added, the authors should explain in the text how the algorithm operates
on that illustration.
An illustration of a specific simplicial complex and the application of the
algorithm to this complex would greatly help readers of this paper. The
algorithm in Section 4.2 is understandable without an illustration, so the
paper is publishable without this change, but such an addition would
strengthen the paper and increase the likelihood of it being read.
--->We have provided a detailed example in the Appendix.

As long as the authors add the restriction on Morse degree to Theorem 1.3 and
the Abstract (noted in a. above), and correct the typo and grammatical errors
below, I believe the paper is ready for publication.

Typo & grammatical errors:
p. 8, line 4: ”maxmima” - Typo. Should be ”maxima”.
--->Fixed.
p. 9, Algorithm setp 3: ”Initialize the join tree to consists...”
Grammatical error. Should be ”to consist”.
--->Fixed.

### 2017-05-03 paint-contour-tree.tex
done with my minor changes

\begin{abstract}
We revisit the classical problem of computing the \emph{contour tree}
of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
The contour tree is a fundamental topological structure that tracks
the evolution of level sets of $f$ and 
has numerous applications in data analysis and visualization.

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
(This is formally incorrect, but gives the right intuition.)
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. We provide formal definitions later. 
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

@@

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
Let the Morse degree of $f$ at any point be at most $3$.
There exists an algorithm to compute $T$
in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
disconnected simplicial complexes.
\end{theorem}

>>>>

\begin{abstract}
We revisit the classical problem of computing the \emph{contour tree}
of a scalar field $f:\MM \to \RR$, where $\MM$ 
is a triangulations of a ball in $\RR^d$.
The contour tree is a fundamental topological structure that tracks
the evolution of level sets of $f$ and 
has numerous applications in data analysis and visualization.

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs. (We assume
that the Morse degree of the function $f$ at any
point is 
at most $3$.)
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes. The contour tree tracks a specific
subset of events, called ``joins" and ``splits".
We provide formal definitions later.
An edge of the contour tree connects two such critical points if one event immediately ``follows"
the other as the sweep plane makes its pass.  
\Fig{relative} and \Fig{sorting} 
show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
and internal nodes are either ``joins" or ``splits".

@@

Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, where $\MM$
is a triangulation of a $d$-dimensional ball.
Let $N$ be the number of facets in $\MM$.
Let the Morse degree of $f$ at any point be at most $3$.
Denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
There exists an algorithm to compute $T$
in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
(Consider the input setting of \Thm{main-corr}.)
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

@@

\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
disconnected simplicial complexes, each of which
is the triangulation of $d$-dimensional ball.
\end{theorem}
